\Large Feature and order manipulations in a free recall task affect memory for current and future lists

### 2023-01-24 paper/main.tex

A key mystery is whether (and how) the sorts of situation models and schemas
that people use to organize their memories of real-world experiences might map
onto the clustering effects that reflect how people organize their memories for
word lists. On one hand, both situation models and clustering effects reflect
statistical regularities in ongoing experiences. Our memory systems exploit
these regularities when generating inferences about the unobserved past and
yet-to-be-experienced future~\citep{XuEtal23, SchaTurk15, RangRitc12,
BoweEtal79, MomeEtal17}. On the other hand, the rich structure of real-world
experiences and other naturalistic stimuli that enable people to form deep and
meaningful situation models and schemas have no obvious analog in simple word
lists. Often, lists in free recall studies are explicitly \textit{designed} to
be devoid of exploitable temporal structure, for example, by sorting the words
in a random order~\citep{Kaha12}.

% feature-rich free recall, basic manipulation conditions, preview of findings
We designed an experimental paradigm to explore how people organize their
memories for simple stimuli (word lists) whose temporal properties change
across different ``situations,'' analogous to how the content of real-world
experiences change across different real-world situations. We asked
participants to study and freely recall a series of word lists
(Fig.~\ref{fig:exp}). In the different conditions in our experiment, we varied
the lists' appearances and presentation orders in different ways. The studied
items (words) were designed to vary along three general dimensions: semantic
(word \textit{category}, and physical \textit{size} of the referent),
lexicographic (word \textit{length} and \textit{first letter}), and visual
(font \textit{color} and the onscreen \textit{location} of each word). We used
two control conditions as a baseline; in these control conditions all of the
lists were sorted randomly, but we manipulated the presence or absence of the
visual features. In two conditions, we manipulated whether the words'
appearances were fixed or variable within each list. In six manipulation
conditions, we asked participants to first study and recall eight lists whose
items were sorted by a target feature (e.g., word category), and then study and
recall an additional eight lists whose items had the same features, but that
were sorted in a random temporal order. We were interested in how these
manipulations affected participants' recall behaviors on early (manipulated)
lists, as well as how order manipulations on early lists affected recall
behaviors on later (randomly ordered) lists. Finally, in an \textit{adaptive}
experimental condition we used participants' recall behaviors on early lists to
manipulate, in real-time, the presentation orders of subsequent lists. In this
adaptive condition, we varied the agreement between how participants preferred
to organize their memories of the studied items versus the orders in which the
items were presented.

@@

We enrolled a total of 491 members of the Dartmouth College community across 11
experimental conditions. The conditions included two controls (feature rich,
reduced), two visual manipulation conditions [reduced (early) and reduced
(late)], six order manipulation conditions (category, size, length, first
letter, color, and location), and a final adaptive condition. Each of these
conditions is described in the \textit{Experimental design} subsection below.

@@

We dropped from our dataset the one participant who reported having abnormal
color vision, as well as 38 participants whose data were corrupted due to
technical failures while running the experiment or during the daily database
merges. In total, this left usable data from 452 participants, broken down by
experimental condition as follows: feature rich (67 participants), reduced (61
participants), reduced (early), (42 participants), reduced (late) (41
participants), category (30 participants), size (30 participants), length (30
participants), first letter (30 participants), color (31 participants),
location (30 participants), and adaptive (60 participants). The participant who
declined to fill out their demographic survey participated in the location
condition, and we verified verbally that they had normal color vision and no
significant reading impairments.

@@

We define each participant's \textit{memory fingerprint} as the set of their
permutation-corrected clustering scores across all dimensions we tracked in our
study, including their six feature-based clustering scores (category, size,
length, first letter, color, and location) and their temporal clustering score.
Conceptually, a participant's memory fingerprint describes their tendency to
order in their recall sequences (and, presumably, organize in memory) the
studied words along each dimension. To obtain stable estimates of these
fingerprints for each participant, we averaged clustering scores across lists.
We also tracked and characterized how participants' fingerprints changed across
lists (e.g.,
Figs.~\ref{fig:fingerprint-trajectories},~\fingerprintTrajectoryRandom).

@@

The presentation orders of some lists in the adaptive condition of our
experiment (see \textit{Adaptive condition}) were sorted according to
participants' \textit{current} memory fingerprint, estimated using all of the
lists they had studied up to that point in the experiment. Because our
experiment incorporated a speech-to-text component, all of the behavioral data
for each participant could be analyzed just a few seconds after the conclusion
of the recall intervals for each list. We used the \texttt{Quail} Python
package~\citep{HeusEtal17} to apply speech-to-text algorithms to the
just-collected data, aggregate the data for the given participant, and estimate
the participant's memory fingerprint using all of their available data up to
that point in the experiment. Two aspects of our implementation are worth
noting. First, because memory fingerprints are computed independently for each
list and then averaged across lists, the already-computed memory fingerprints
for earlier lists could be cached and loaded as needed in future computations.
This meant that our computations pertaining to updating our estimate of a
participant's memory fingerprint only needed to consider data from the most
recent list. Second, each element of the null distributions of uncorrected
fingerprint scores (see \textit{Permutation-corrected feature clustering
scores}) could be estimated independently from the others. This enabled us to
make use of the testing computers' multi-core CPU architectures by considering
(in parallel) elements of the null distributions in batches of eight (i.e., the
number of CPU cores on each testing computer). Taken together, we were able to
compress the relevant computations into just a few seconds of computing time.
The combined processing time for the speech-to-text algorithm, fingerprint
computations, and permutation-based ordering procedure (described next) easily
fit within the inter-list intervals, where participants paused for a self-paced
break before moving on to study and recall the next list.

@@

In the adaptive condition of our experiment, the presentation orders for
\textit{stabilize} and \textit{destabilize} lists were chosen to either
maximally or minimally (respectively) comport with participants' memory
fingerprints. Given a participant's memory fingerprint and a to-be-presented
set of items, we designed a permutation-based procedure for ordering the items.
First, we dropped from the participant's fingerprint the temporal clustering
score. For the remaining feature dimensions, we arranged the clustering scores
in the fingerprint into a template vector, $f$. Second, we computed $n = 2500$
random permutations of the to-be-presented items. These permutations served as
candidate presentation orders. We sought to select the specific order that most
(or least) closely matched $f$. Third, for each random permutation, we computed
the (permutation-corrected) ``fingerprint,'' treating the permutation as though
it were a potential ``perfect'' recall sequence. (We did not include temporal
clustering scores in these fingerprints.) This yielded a ``simulated
fingerprint'' vector, $\hat{f}_p$ for each permutation $p$. We used these
simulated fingerprints to select a specific permutation, $i$, that either
maximized (for stabilize lists) or minimized (for destabilize lists) the
correlation between $\hat{f}_i$ and $f$.

@@

With these differences in mind, we next compared participants' memories on
early versus late lists for two additional experimental conditions (see
\textit{Random conditions}, Fig.~\dynamicsRandom). In a \textit{reduced
(early)} condition, we held the irrelevant visual features constant on early
lists, but allowed them to vary randomly on late lists. In a \textit{reduced
(late)} condition, we allowed the irrelevant visual features to vary randomly
on early lists, but held them constant on late lists. Given our above findings
that (a) participants tended to remember more words and exhibit stronger
clustering effects on feature rich (versus reduced) lists, and (b) participants
tended to remember more words and exhibit stronger clustering effects on early
(versus late) lists, we expected these early versus late differences to be
enhanced in the reduced (early) condition and diminished in the reduced (late)
condition. However, to our surprise, participants in \textit{neither} condition
exhibited reliable early versus late differences in accuracy (reduced (early):
$t(41) = 1.499, p = 0.141$; reduced (late): $t(40) = 1.462, p = 0.152$),
temporal clustering (reduced (early): $t(41) = 0.998, p = 0.324$; reduced
(late): $t(40) = 1.099, p = 0.278$), nor feature-based clustering (reduced
(early), category: $t(41) = 0.753, p = 0.456$; reduced (early), size: $t(41) =
0.721, p = 0.475$; reduced (early), length: $t(41) = 0.493, p = 0.625$; reduced
(early), first letter: $t(41) = 0.780, p = 0.440$; reduced (late), category:
$t(40) = -0.086, p = 0.932$; reduced (late), size: $t(40) = 0.746, p = 0.460$;
reduced (late), length: $t(40) = 1.476, p = 0.148$; reduced (late), first
letter: $t(40) = 0.966, p = 0.340$). We hypothesized that adding or removing
the irrelevant features was acting as a sort of ``event boundary'' between
early and late lists. In prior work, we (and others) have found that memories
formed just after event boundaries can be enhanced~\citep[e.g., due to less
contextual interference between pre- and post-boundary items;][]{MannEtal16,
PettEtal16, GoldEtal17, FlorEtal17}.

We found that \textit{adding} irrelevant visual features on later lists that
had not been present on early lists (as in the reduced (early) condition)
served to enhance recall performance relative to conditions where all lists had
the same blends of features (accuracy for feature rich versus reduced (early):
$t(107) = -2.230, p = 0.028$; reduced versus reduced (early): $t(101) = -2.045,
p = 0.043$; also see Fig.~\accuracyByList A). However, \textit{subtracting}
irrelevant visual features on later lists that \textit{had} been present on
early lists (as in the reduced (late) condition) did not appear to impact
recall performance (accuracy for feature rich versus reduced (late): $t(106) =
-0.638, p = 0.525$; reduced versus reduced (late): $t(100) = -0.407, p =
0.685$). These comparisons suggest that recall accuracy has a directional
component: accuracy is affected differently by removing features later
that had been present earlier versus adding features later that had
\textit{not} been present earlier. In contrast, we found that participants
exhibited more temporal and feature-based clustering when we added irrelevant
visual features to \textit{any} lists (comparisons of clustering on feature
rich versus reduced lists are reported above; temporal clustering in reduced
versus reduced (early) and reduced versus reduced (late) conditions: $t$s $\leq
-9.780$, $p$s $< 0.001$; feature-based clustering in reduced versus reduced
(early) and reduced versus reduced (late) conditions: $t$s $\leq -5.443$, $p$s
$< 0.001$). Temporal and feature-based clustering were not reliably different
in the feature rich, reduced (early), and reduced (late) conditions (temporal
clustering in feature rich versus reduced (early) and feature rich versus
reduced (late) conditions: $t$s $\geq -1.434$, $p$s $\geq 0.154$; feature-based
clustering in feature rich versus reduced (early) and feature rich versus
reduced (late) conditions: $t$s $\geq -1.359$, $p$s $> 0.177$).

Taken together, our findings thus far suggest that adding item features that
change over time, even when they vary randomly and independently of the items,
can enhance participants' overall memory performance and can also enhance
temporal and feature-based clustering. To the extent that the number of item
features that vary from moment to moment approximates the ``richness'' of
participants' experiences, our findings suggest that participants remember
``richer'' stimuli better and organize richer stimuli more reliably in their
memories. Next, we turn to examine the memory effects of varying the temporal
ordering of different stimulus features while holding the features themselves
constant. We hypothesized that changing the orders in which participants were
exposed to the words on a given list might enhance (or diminish) the relative
influence of different features. For example, presenting a set of words
alphabetically might enhance participants' attention to the studied items'
first letters, whereas sorting the same list of words by semantic category
might instead enhance participants' attention to the words' semantic
attributes. Importantly, we expected these order manipulations to hold even
when the variation in the total set of features (across words) was held
constant across lists (e.g., unlike in the reduced (early) and reduced (late)
conditions, where variations in visual features were added or removed from a
subset of the lists participants studied).

@@

When we closely examined the sequences of words participants recalled from
early order-manipulated lists (Fig.~\ref{fig:recall-dynamics}A, top panel), we
noticed several differences from the dynamics of participants' recalls of
randomly ordered lists (Figs.~\dynamicsRandom,~\recallInit). One difference is
that participants in the category condition (dark purple curves,
Fig.~\ref{fig:recall-dynamics}) most often initiated recall with the
fourth-from-last item (\textit{Recall initiation}, top left panel), whereas
participants who recalled randomly ordered lists tended to initiate recall with
either the first or last list items (Fig.~\dynamicsRandom, top left panel). We
hypothesized that the participants might be ``clumping'' their recalls into
groups of items that shared category labels. Indeed, when we compared the
positions of feature changes in the study sequence
(Fig.~\ref{fig:recall-dynamics}B; see \textit{Identifying event boundaries})
with the positions of items participants recalled first, we noticed a striking
correspondence in both semantic conditions. Specifically, on category-ordered
lists, the category labels changed every four items on average (dark purple
peaks in Fig.~\ref{fig:recall-dynamics}B), and participants also seemed to
display an increased tendency (relative to other order manipulation and random
conditions) to initiate recall of category-ordered lists with items whose study
positions were integer multiples of four. Similarly, for size-ordered lists,
the size labels changed every eight items on average (light purple peaks in
Fig.~\ref{fig:recall-dynamics}B), and participants also seemed to display an
increased tendency to initiate recall of size-ordered lists with items whose
study positions were integer multiples of eight. A second striking difference
is that participants in the category condition exhibited a much steeper lag-CRP
(Fig.~\ref{fig:recall-dynamics}A, top middle panel) than participants in other
conditions. (This is another expression of participants' increased tendencies
to temporally cluster their recalls on category-ordered lists, as we reported
above.) Taken together, these order-specific idiosyncrasies suggest a
hierarchical set of influences on participants' memories. At longer timescales,
``event boundaries'' (to use the term loosely) can be induced across lists by
adding or removing irrelevant visual features. At shorter timescales, ``event
boundaries'' can be induced across items (within a single list) by adjusting
how item features change throughout the list.

The above comparisons between memory performance on early lists in the order
manipulation versus feature rich conditions highlight how sorted lists are
remembered differently from random lists. We also wondered how sorting lists
along each feature dimension influenced memory relative to sorting lists along
the other feature dimensions. Participants trended towards remembering early
lists that were sorted semantically better than lexicographically sorted lists
($t(118) = 1.936, p = 0.055$). Participants also remembered visually sorted
lists better than lexicographically sorted lists ($t(119) = 2.145, p = 0.034$).
However, participants showed no reliable differences in recall for semantically
versus visually sorted lists ($t(119) = 0.113, p = 0.910$). Participants
temporally clustered semantically sorted lists more strongly than either
lexicographically ($t(118) = 5.572, p < 0.001$) or visually ($t(119) = 6.215, p
< 0.001$) sorted lists, but did not show reliable differences in temporal
clustering on lexicographically versus visually sorted lists ($t(119) = 0.189,
p = 0.850$). Participants also showed reliably more semantic clustering on
semantically sorted lists than lexicographically (category: $t(118) = 3.492, p
= 0.001$, size: $t(118) = 3.972, p < 0.001$) or visually (category: $t(119) =
2.702, p = 0.008$, size: $t(119) = 4.230, p < 0.001$) sorted lists; more
lexicographic clustering on lexicographically sorted lists than semantically
(length: $t(118) = 3.112, p = 0.002$; first letter: $t(118) = 3.686, p =
0.000$) or visually (length: $t(119) = 3.024, p = 0.003$; first letter: $t(119)
= 2.644, p = 0.009$) sorted lists; and more visual clustering on visually
sorted lists than semantically (color: $t(119) = -2.659, p = 0.009$; location:
$t(119) = -4.604, p < 0.001$) or lexicographically (color: $t(119) = -2.366, p
= 0.020$; location: $t(119) = -4.265, p < 0.001$) sorted lists. In summary,
sorting lists by different features appeared to have slightly different effects
on overall memory performance and temporal clustering.  Participants also tended to
cluster their recalls along a given feature dimension more when the studied
lists were (versus were not) sorted along that dimension.

@@

We also looked for more subtle group-level patterns. For example, perhaps
sorting early lists by one feature dimension could affect how participants
cluster \textit{other} features (on early and/or late lists) as well. We
defined participants' \textit{memory fingerprints} as the set of their temporal
and feature clustering scores (see \textit{Memory fingerprints}). A
participant's memory fingerprint describes how they tend to retrieve memories
of the studied items, perhaps searching through several feature spaces (or
along several representational dimensions). To gain insights into the dynamics
of how participants' clustering scores tended to change over time, we computed
the average (across participants) fingerprint from each list, from each order
manipulation condition (Fig.~\ref{fig:fingerprint-trajectories}). We projected
these fingerprints into a two-dimensional space to help visualize the dynamics
(top panels; see \textit{Computing low-dimensional embeddings of memory
fingerprints}). We found that participants' average fingerprints tended to
remain relatively stable on early lists, and exhibited a ``jump'' to another
stable state on later lists. The sizes of these jumps varied somewhat across
conditions (the Euclidean distances between fingerprints in their original high
dimensional spaces are displayed in the bottom panels). We also averaged the
fingerprints across early and late lists, respectively, for each condition
(Fig.~\ref{fig:fingerprint-trajectories}B). We found that participants'
fingerprints on early lists seem to be influenced by the order manipulations
for those lists (see the locations of the circles in
Fig.~\ref{fig:fingerprint-trajectories}B). There also seemed to be some
consistency across different features within a broader type. For example, both
semantic feature conditions (category and size; purple markers) diverge in a
similar direction from the group; both lexicographic feature conditions (length
and first letter; yellow markers) diverge in a similar direction; and both
visual conditions (color and location; green) also diverge in a similar
direction. But on late lists, participants' fingerprints seem to return to a
common state that is roughly shared across conditions (i.e., the stars in that
panel are clumped together).

@@

When we held the within-list variability in participants' visual experiences
fixed across lists (in the feature rich and reduced conditions), they
remembered more words from early lists than from late lists. For feature rich
lists, they also showed stronger clustering for early versus late lists.
However, when we \textit{varied} participants' visual experiences across lists
(in the ``reduced (early)'' and ``reduced (late)'' conditions), these early
versus late accuracy and clustering differences disappeared. Abruptly changing
how irrelevant visual features varied across words seemed to act as a sort of
``event boundary'' that partially reset how participants processed and
remembered post-boundary lists. Within-list clustering also increased in these
manipulations, suggesting that the ``within-event'' words were being more tightly
associated with each other.

>>>>

A key mystery is whether (and how) the sorts of situation models and schemas
that people use to organize their memories of real-world experiences might map
onto the clustering effects that reflect how people organize their memories for
word lists. On one hand, both situation models and clustering effects reflect
statistical regularities in ongoing experiences. Our memory systems exploit
these regularities when generating inferences about the unobserved past and
yet-to-be-experienced future~\citep{XuEtal23, SchaTurk15, RangRitc12,
BoweEtal79, MomeEtal17}. On the other hand, the rich structures of real-world
experiences and other naturalistic stimuli that enable people to form deep and
meaningful situation models and schemas have no obvious analogs in simple word
lists. Often, lists in free recall studies are explicitly \textit{designed} to
be devoid of exploitable temporal structure, for example, by sorting the words
in a random order~\citep{Kaha12}.

% feature-rich free recall, basic manipulation conditions, preview of findings
We designed an experimental paradigm to explore how people organize their
memories for simple stimuli (word lists) whose temporal properties change
across different ``situations,'' analogous to how the content of real-world
experiences change across different real-world situations. We asked
participants to study and freely recall a series of word lists
(Fig.~\ref{fig:exp}). In the different conditions in our experiment, we varied
the lists' appearances and presentation orders in different ways. The studied
items (words) were designed to vary along three general dimensions: semantic
(word \textit{category} and physical \textit{size} of the referent),
lexicographic (word \textit{length} and \textit{first letter}), and visual
(font \textit{color} and the onscreen \textit{location} of each word). We used
two control conditions as a baseline; in these control conditions all of the
lists were sorted randomly, but we manipulated the presence or absence of the
visual features. In two conditions, we manipulated whether the words'
appearances were fixed or variable within each list. In six conditions, we
asked participants to first study and recall eight lists whose items were
sorted by a target feature (e.g., word category), and then study and recall an
additional eight lists whose items had the same features, but that were sorted
in a random temporal order. We were interested in how these manipulations
affected participants' recall behaviors on early (manipulated) lists, as well
as how order manipulations on early lists affected recall behaviors on later
(randomly ordered) lists. Finally, in an \textit{adaptive} experimental
condition we used participants' recall behaviors on early lists to manipulate,
in real-time, the presentation orders of subsequent lists. In this adaptive
condition, we varied the agreement between how participants preferred to
organize their memories of the studied items versus the orders in which the
items were presented.

@@

We enrolled a total of 491 members of the Dartmouth College community across 11
experimental conditions. The conditions included two controls (feature rich and
reduced), two visual manipulation conditions [reduced (early) and reduced
(late)], six order manipulation conditions (category, size, length, first
letter, color, and location), and a final adaptive condition. Each of these
conditions is described in the \textit{Experimental design} subsection below.

@@

We dropped from our dataset the one participant who reported having abnormal
color vision, as well as 38 participants whose data were corrupted due to
technical failures while running the experiment or during the daily database
merges. In total, this left usable data from 452 participants, broken down by
experimental condition as follows: feature rich (67 participants), reduced (61
participants), reduced (early) (42 participants), reduced (late) (41
participants), category (30 participants), size (30 participants), length (30
participants), first letter (30 participants), color (31 participants),
location (30 participants), and adaptive (60 participants). The participant who
declined to fill out their demographic survey participated in the location
condition, and we verified verbally that they had normal color vision and no
significant reading impairments.

@@

We define each participant's \textit{memory fingerprint} as the set of their
permutation-corrected clustering scores across all dimensions we tracked in our
study, including their six feature-based clustering scores (category, size,
length, first letter, color, and location) and their temporal clustering score.
Conceptually, a participant's memory fingerprint describes their tendency to
order in their recall sequences (and, presumably, organize in memory) the
studied words along each dimension. To obtain stable estimates of these
fingerprints for each participant, we averaged their clustering scores across
lists. We also tracked and characterized how participants' fingerprints changed
across lists (e.g.,
Figs.~\ref{fig:fingerprint-trajectories},~\fingerprintTrajectoryRandom).

@@

The presentation orders of some lists in the adaptive condition of our
experiment (see \textit{Adaptive condition}) were sorted according to
participants' \textit{current} memory fingerprint, estimated using all of the
lists they had studied up to that point in the experiment. Because our
experiment incorporated a speech-to-text component, all of the behavioral data
for each participant could be analyzed just a few seconds after the conclusion
of the recall intervals for each list. We used the \texttt{Quail} Python
package~\citep{HeusEtal17} to apply speech-to-text algorithms to the
just-collected audio data, aggregate the data for the given participant, and estimate
the participant's memory fingerprint using all of their available data up to
that point in the experiment. Two aspects of our implementation are worth
noting. First, because memory fingerprints are computed independently for each
list and then averaged across lists, the already-computed memory fingerprints
for earlier lists could be cached and loaded as needed in future computations.
This meant that our computations pertaining to updating our estimate of a
participant's memory fingerprint only needed to consider data from the most
recent list. Second, each element of the null distributions of uncorrected
fingerprint scores (see \textit{Permutation-corrected feature clustering
scores}) could be estimated independently from the others. This enabled us to
make use of the testing computers' multi-core CPU architectures by considering
(in parallel) elements of the null distributions in batches of eight (i.e., the
number of CPU cores on each testing computer). Taken together, we were able to
compress the relevant computations into just a few seconds of computing time.
The combined processing time for the speech-to-text algorithm, fingerprint
computations, and permutation-based ordering procedure (described next) easily
fit within the inter-list intervals, where participants paused for a self-paced
break before moving on to study and recall the next list.

@@

In the adaptive condition of our experiment, the presentation orders for
\textit{stabilize} and \textit{destabilize} lists were chosen to either
maximally or minimally (respectively) comport with participants' memory
fingerprints. Given a participant's memory fingerprint and a to-be-presented
set of items, we designed a permutation-based procedure for ordering the items.
First, we dropped from the participant's fingerprint the temporal clustering
score. For the remaining feature dimensions, we arranged the clustering scores
in the fingerprint into a template vector, $f$. Second, we computed $n = 2500$
random permutations of the to-be-presented items. These permutations served as
candidate presentation orders. We sought to select the specific order that most
(or least) closely matched $f$. Third, for each random permutation, we computed
the (permutation-corrected) ``fingerprint,'' treating the permutation as though
it were a potential ``perfect'' recall sequence. (We did not include temporal
clustering scores in these fingerprints, since the temporal clustering score
for every permutation is always equal to 1.) This yielded a ``simulated
fingerprint'' vector, $\hat{f}_p$ for each permutation $p$. We used these
simulated fingerprints to select a specific permutation, $i$, that either
maximized (for stabilize lists) or minimized (for destabilize lists) the
correlation between $\hat{f}_i$ and $f$.

@@

With these differences in mind, we next compared participants' memories on
early versus late lists for two additional experimental conditions (see
\textit{Random conditions}, Fig.~\dynamicsRandom). In a \textit{reduced
(early)} condition, we held the visual features constant on early lists, but
allowed them to vary randomly on late lists. In a \textit{reduced (late)}
condition, we allowed the visual features to vary randomly on early lists, but
held them constant on late lists. Given our above findings that (a)
participants tended to remember more words and exhibit stronger clustering
effects on feature rich (versus reduced) lists, and (b) participants tended to
remember more words and exhibit stronger clustering effects on early (versus
late) lists, we expected these early versus late differences to be enhanced in
the reduced (early) condition and diminished in the reduced (late) condition.
However, to our surprise, participants in \textit{neither} condition exhibited
reliable early versus late differences in accuracy (reduced (early): $t(41) =
1.499, p = 0.141$; reduced (late): $t(40) = 1.462, p = 0.152$), temporal
clustering (reduced (early): $t(41) = 0.998, p = 0.324$; reduced (late): $t(40)
= 1.099, p = 0.278$), nor feature-based clustering (reduced (early), category:
$t(41) = 0.753, p = 0.456$; reduced (early), size: $t(41) = 0.721, p = 0.475$;
reduced (early), length: $t(41) = 0.493, p = 0.625$; reduced (early), first
letter: $t(41) = 0.780, p = 0.440$; reduced (late), category: $t(40) = -0.086,
p = 0.932$; reduced (late), size: $t(40) = 0.746, p = 0.460$; reduced (late),
length: $t(40) = 1.476, p = 0.148$; reduced (late), first letter: $t(40) =
0.966, p = 0.340$). We hypothesized that adding or removing the variability in
the visual features was acting as a sort of ``event boundary'' between early
and late lists. In prior work, we (and others) have found that memories formed
just after event boundaries can be enhanced~\citep[e.g., due to less contextual
interference between pre- and post-boundary items;][]{MannEtal16, PettEtal16,
GoldEtal17, FlorEtal17}.

We found that \textit{adding} incidental visual features on later lists that
had not been present on early lists (as in the reduced (early) condition)
served to enhance recall performance relative to conditions where all lists had
the same blends of features (accuracy for feature rich versus reduced (early):
$t(107) = -2.230, p = 0.028$; reduced versus reduced (early): $t(101) = -2.045,
p = 0.043$; also see Fig.~\accuracyByList A). However, \textit{subtracting}
irrelevant visual features on later lists that \textit{had} been present on
early lists (as in the reduced (late) condition) did not appear to impact
recall performance (accuracy for feature rich versus reduced (late): $t(106) =
-0.638, p = 0.525$; reduced versus reduced (late): $t(100) = -0.407, p =
0.685$). These comparisons suggest that recall accuracy has a directional
component: accuracy is affected differently by removing features later
that had been present earlier versus adding features later that had
\textit{not} been present earlier. In contrast, we found that participants
exhibited more temporal and feature-based clustering when we added incidental
visual features to \textit{any} lists (comparisons of clustering on feature
rich versus reduced lists are reported above; temporal clustering in reduced
versus reduced (early) and reduced versus reduced (late) conditions: $t$s $\leq
-9.780$, $p$s $< 0.001$; feature-based clustering in reduced versus reduced
(early) and reduced versus reduced (late) conditions: $t$s $\leq -5.443$, $p$s
$< 0.001$). Temporal and feature-based clustering were not reliably different
in the feature rich, reduced (early), and reduced (late) conditions (temporal
clustering in feature rich versus reduced (early) and feature rich versus
reduced (late) conditions: $t$s $\geq -1.434$, $p$s $\geq 0.154$; feature-based
clustering in feature rich versus reduced (early) and feature rich versus
reduced (late) conditions: $t$s $\geq -1.359$, $p$s $> 0.177$).

Taken together, our findings thus far suggest that adding item features that
change over time, even when they vary randomly and independently of the items,
can enhance participants' overall memory performance and can also enhance
temporal and feature-based clustering. To the extent that the number of item
features that vary from moment to moment approximates the ``richness'' of
participants' experiences, our findings suggest that participants remember
``richer'' stimuli better and organize richer stimuli more reliably in their
memories. Next, we turn to examine the memory effects of varying the temporal
ordering of different stimulus features. We hypothesized that changing the
orders in which participants were exposed to the words on a given list might
enhance (or diminish) the relative influence of different features. For
example, presenting a set of words alphabetically might enhance participants'
attention to the studied items' first letters, whereas sorting the same list of
words by semantic category might instead enhance participants' attention to the
words' semantic attributes. Importantly, we expected these order manipulations
to hold even when the variation in the total set of features (across words) was
held constant across lists (e.g., unlike in the reduced (early) and reduced
(late) conditions, where variations in visual features were added or removed
from a subset of the lists participants studied).

@@

When we closely examined the sequences of words participants recalled from
early order-manipulated lists (Fig.~\ref{fig:recall-dynamics}A, top panel), we
noticed several differences from the dynamics of participants' recalls of
randomly ordered lists (Figs.~\dynamicsRandom,~\recallInit). One difference is
that participants in the category condition (dark purple curves,
Fig.~\ref{fig:recall-dynamics}) most often initiated recall with the
fourth-from-last item (\textit{Recall initiation}, top left panel), whereas
participants who recalled randomly ordered lists tended to initiate recall with
either the first or last list items (Fig.~\dynamicsRandom, top left panel). We
hypothesized that the participants might be ``clumping'' their recalls into
groups of items that shared category labels. Indeed, when we compared the
positions of feature changes in the study sequence
(Fig.~\ref{fig:recall-dynamics}B; see \textit{Identifying event boundaries})
with the positions of items participants recalled first, we noticed a striking
correspondence in both semantic conditions. Specifically, on category-ordered
lists, the category labels changed every four items on average (dark purple
peaks in Fig.~\ref{fig:recall-dynamics}B), and participants also seemed to
display an increased tendency (relative to other order manipulation and random
conditions) to initiate recall of category-ordered lists with items whose study
positions were integer multiples of four. Similarly, for size-ordered lists,
the size labels changed every eight items on average (light purple peaks in
Fig.~\ref{fig:recall-dynamics}B), and participants also seemed to display an
increased tendency to initiate recall of size-ordered lists with items whose
study positions were integer multiples of eight. A second striking difference
is that participants in the category condition exhibited a much steeper lag-CRP
(Fig.~\ref{fig:recall-dynamics}A, top middle panel) than participants in other
conditions. (This is another expression of participants' increased tendencies
to temporally cluster their recalls on category-ordered lists, as we reported
above.) Taken together, these order-specific idiosyncrasies suggest a
hierarchical set of influences on participants' memories. At longer timescales,
``event boundaries'' (to use the term loosely) can be induced across lists by
adding or removing incidental visual features. At shorter timescales, ``event
boundaries'' can be induced across items (within a single list) by adjusting
how item features change throughout the list.

The above comparisons between memory performance on early lists in the order
manipulation versus feature rich conditions highlight how sorted lists are
remembered differently from random lists. We also wondered how sorting lists
along each feature dimension influenced memory relative to sorting lists along
the other feature dimensions. Participants trended towards remembering early
lists that were sorted semantically better than lexicographically sorted lists
($t(118) = 1.936, p = 0.055$). Participants also remembered visually sorted
lists better than lexicographically sorted lists ($t(119) = 2.145, p = 0.034$).
However, participants showed no reliable differences in recall for semantically
versus visually sorted lists ($t(119) = 0.113, p = 0.910$). Participants
temporally clustered semantically sorted lists more strongly than either
lexicographically ($t(118) = 5.572, p < 0.001$) or visually ($t(119) = 6.215, p
< 0.001$) sorted lists, but did not show reliable differences in temporal
clustering on lexicographically versus visually sorted lists ($t(119) = 0.189,
p = 0.850$). Participants also showed reliably more semantic clustering on
semantically sorted lists than lexicographically (category: $t(118) = 3.492, p
= 0.001$, size: $t(118) = 3.972, p < 0.001$) or visually (category: $t(119) =
2.702, p = 0.008$, size: $t(119) = 4.230, p < 0.001$) sorted lists; more
lexicographic clustering on lexicographically sorted lists than semantically
(length: $t(118) = 3.112, p = 0.002$; first letter: $t(118) = 3.686, p <
0.001$) or visually (length: $t(119) = 3.024, p = 0.003$; first letter: $t(119)
= 2.644, p = 0.009$) sorted lists; and more visual clustering on visually
sorted lists than semantically (color: $t(119) = -2.659, p = 0.009$; location:
$t(119) = -4.604, p < 0.001$) or lexicographically (color: $t(119) = -2.366, p
= 0.020$; location: $t(119) = -4.265, p < 0.001$) sorted lists. In summary,
sorting lists by different features appeared to have slightly different effects
on overall memory performance and temporal clustering.  Participants also tended to
cluster their recalls along a given feature dimension more when the studied
lists were (versus were not) sorted along that dimension.

@@

We also looked for more subtle group-level patterns. For example, perhaps
sorting early lists by one feature dimension could affect how participants
cluster \textit{other} features (on early and/or late lists) as well. We
defined participants' \textit{memory fingerprints} as the set of their temporal
and feature clustering scores (see \textit{Memory fingerprints}). A
participant's memory fingerprint describes how they tend to retrieve memories
of the studied items, perhaps searching in parallel through several feature
spaces (or along several representational dimensions). To gain insights into
the dynamics of how participants' clustering scores tended to change over time,
we computed the average (across participants) fingerprint from each list, from
each order manipulation condition (Fig.~\ref{fig:fingerprint-trajectories}). We
projected these fingerprints into a two-dimensional space to help visualize the
dynamics (top panels; see \textit{Computing low-dimensional embeddings of
memory fingerprints}). We found that participants' average fingerprints tended
to remain relatively stable on early lists, and exhibited a ``jump'' to another
stable state on later lists. The sizes of these jumps varied somewhat across
conditions (the Euclidean distances between fingerprints in their original high
dimensional spaces are displayed in the bottom panels). We also averaged the
fingerprints across early and late lists, respectively, for each condition
(Fig.~\ref{fig:fingerprint-trajectories}B). We found that participants'
fingerprints on early lists seem to be influenced by the order manipulations
for those lists (see the locations of the circles in
Fig.~\ref{fig:fingerprint-trajectories}B). There also seemed to be some
consistency across different features within a broader type. For example, both
semantic feature conditions (category and size; purple markers) diverge in a
similar direction from the group; both lexicographic feature conditions (length
and first letter; yellow markers) diverge in a similar direction; and both
visual conditions (color and location; green) also diverge in a similar
direction. But on late lists, participants' fingerprints seem to return to a
common state that is roughly shared across conditions (i.e., the stars in that
panel are clumped together).

@@

When we held the within-list variability in participants' visual experiences
fixed across lists (in the feature rich and reduced conditions), they
remembered more words from early lists than from late lists. For feature rich
lists, they also showed stronger clustering for early versus late lists.
However, when we \textit{varied} participants' visual experiences across lists
(in the ``reduced (early)'' and ``reduced (late)'' conditions), these early
versus late accuracy and clustering differences disappeared. Abruptly changing
how incidental visual features varied across words seemed to act as a sort of
``event boundary'' that partially reset how participants processed and
remembered post-boundary lists. Within-list clustering also increased in these
manipulations, suggesting that the ``within-event'' words were being more tightly
associated with each other.
